Supplement to ``Local clustering of density upper level sets''

### 2019-01-16 icml/icml_local_density_clustering_supplement.tex
added mixing time estimate lemmas

%%% Vectors
\newcommand{\pbf}{\mathbf{p}}
\newcommand{\ebf}[1]{\mathbf{e}_{#1}}
\newcommand{\pibf}{\bm{\pi}}

@@

%%% Probability distributions (and related items)
\newcommand{\Pbb}{\mathbb{P}}
\newcommand{\Cbb}{\mathbb{C}}
\newcommand{\Ebb}{\mathbb{E}}

%%% Sets
\newcommand{\Cset}{\mathcal{C}}
\newcommand{\Aset}{\mathcal{A}}
\newcommand{\Asig}{\Aset_{\sigma}}
\newcommand{\Csig}{\Cset_{\sigma}}
\newcommand{\Asigr}{\Aset_{\sigma,\sigma + r}}
\newcommand{\Csigr}{\Cset_{\sigma,\sigma + r}}

@@

Throughout, we will fix $\Aset \subset \Rd$ to be an arbitrary set. To simplify expressions, for the $\sigma$-expansion $\Asig$, we will write the set difference between $\Asig$ and the $(\sigma + r)$-expansion $\Aset_{\sigma + r}$ as 
\begin{equation*}
\Asigr := \set{x: 0 < \rho(x, \Asig) \leq r},
\end{equation*}
where $\rho(x, \Aset) = \min_{x' \in \Aset} \norm{x - x'}$.
.

@@

\subsection{Cut and volume estimates}
\begin{lemma}
	\label{lem: expected_number_boundary_points}
	Under the setup and conditions of Theorem \ref{thm: conductance_upper_bound}, and for any $r < \sigma/2d$,
	\begin{equation*}
	\Pbb(\Csigr) \leq 2 \nu(\Csig) \frac{rd}{\sigma}  \left(\lambda_{\sigma} - \frac{r^{\gamma}}{\gamma + 1}\right)
	\end{equation*}	
\end{lemma}
\begin{proof}
	Recalling that $f$ is the density function for $\Pbb$, we have
	\begin{equation}
	\label{eqn: integral_over_epsilon_neighborhood}
	\Pbb(\Csigr) = \int_{\Csigr} f(x) dx
	\end{equation}
	We partition $\Csigr$ into slices, based on distance from $\Csig$, as follows: for $k \in \N$,
	\begin{equation*}
	\mathcal{T}_{i,k} = \set{x \in \Rd: t_{i,k} < \frac{\rho(x, \Csig)}{r} \leq t_{i+1,k}}, ~~ \Csigr = \bigcup_{i = 0}^{k-1} \mathcal{T}_{i,k}
	\end{equation*}
	where $t_i = i/k$ for $i = 0, \ldots, k - 1$. As a result,
	\begin{equation*}
	\int_{\Csigr} f(x) dx = \sum_{i = 0}^{k-1} \int_{\mathcal{T}_{i,k}} f(x) dx \leq \sum_{i = 0}^{k-1} \nu(\mathcal{T}_{i,k}) \max_{x \in \mathcal{T}_{i,k}} f(x).
	\end{equation*}
	We substitute
	\begin{equation*}
	\nu(\mathcal{T}_{i,k}) = \nu(\Csig + rt_{i+1,k}B) - \nu(\Csig + rt_{i,k}B) := \nu_{i+1,k} - \nu_{i,k}. 
	\end{equation*}
	where for simplicity we've written $\nu_{i,k} = \nu(\Csig + rt_{i+1,k}B)$.
	This, in concert with the upper bound
	\begin{equation*}
	\max_{x \in \mathcal{T}_{i,k}} f(x) \leq \lambda_{\sigma} - (rt_{i,k})^{\gamma},
	\end{equation*}
	which follows from \ref{asmp: bounded_density} and \ref{asmp: low_noise_density}, yields
	\begin{align}
	\label{eqn: telescoping_sum}
	\sum_{i = 0}^{k-1} \nu(\mathcal{T}_{i,k}) \max_{x \in \mathcal{T}_{i,k}} f(x) & \leq \sum_{i = 0}^{k-1} \biggl\{ \nu_{i+1,k} - \nu_{i,k} \biggr\} \biggl( \lambda_{\sigma} - (rt_{i,k})^{\gamma} \biggr) \nonumber \\
	& = \sum_{i = 1}^{k} 
	\underbrace{\nu_{i,k} \biggl( \left[\lambda_{\sigma} - (rt_{i,k})^{\gamma}\right] -  \left[\lambda_{\sigma} - (rt_{i-1,k})^{\gamma}\right]\biggr)}_{:= \Sigma_k} + \underbrace{\biggl(\nu_{k,k}\left[\lambda_{\sigma} - r^{\gamma}\right] - \nu_{1,k}\lambda_{\sigma} \biggr)}_{:= \xi_k}
	\end{align}

@@

>>>>

%%% Vectors
\newcommand{\pbf}{\mathbf{p}}
\newcommand{\ebf}[1]{\mathbf{e}_{#1}}
\newcommand{\pibf}{\bm{\pi}}
\newcommand{\rhobf}{\bm{\rho}}

@@

%%% Probability distributions (and related items)
\newcommand{\Pbb}{\mathbb{P}}
\newcommand{\Qbb}{\mathbb{Q}}
\newcommand{\Cbb}{\mathbb{C}}
\newcommand{\Ebb}{\mathbb{E}}

%%% Sets
\newcommand{\Sset}{\mathcal{S}}
\newcommand{\Cset}{\mathcal{C}}
\newcommand{\Aset}{\mathcal{A}}
\newcommand{\Asig}{\Aset_{\sigma}}
\newcommand{\Csig}{\Cset_{\sigma}}
\newcommand{\Asigr}{\Aset_{\sigma,\sigma + r}}
\newcommand{\Csigr}{\Cset_{\sigma,\sigma + r}}

@@

Throughout, we will fix $\Aset \subset \Rd$ to be an arbitrary set. To simplify expressions, for the $\sigma$-expansion $\Asig$, we will write the set difference between $\Asig$ and the $(\sigma + r)$-expansion $\Aset_{\sigma + r}$ as 
\begin{equation*}
\Asigr := \set{x: 0 < \rho(x, \Asig) \leq r},
\end{equation*}
where $\rho(x, \Aset) = \min_{x' \in \Aset} \norm{x - x'}$.

@@

As part of the proof of Theorem \ref{thm: inverse_mixing_time_lower_bound}, we will require an estimate of a function $g(t)$ for $t \in [x_0, 1/2]$ for some $x_0 > 0$. For $m > 0$ and $x_0 = t_0 < t_1 < \ldots < t_m = 1/2$, define the \emph{stepwise approximation to $g$} to be $\bar{g}$, given by 
\begin{equation}
\label{eqn: stepwise_approximation}
\bar{g}(t) = g(t_i), ~~ \text{ for $t \in [t_{i-1}, t_i]$ }
\end{equation}
\begin{lemma}
	\label{lem: stepwise_approximation}
	Fix
	\begin{equation*}
	g(t) = \log\left(\frac{1}{t}\right) \text{ for $x \in [x_0, 1/2]$}
	\end{equation*}
	 If for all $i$ in $1,\ldots,m$, $(t_i - t_{i - 1}) \leq x_0/2$, then $g(t) \geq \bar{g}(t)\geq g(t) / 2$.
\end{lemma}
\begin{proof}
	The upper bound $g(t) \geq \bar{g}(t)$ follows immediately from the fact that $g(t)$ is a decreasing function.

	By the concavity of the $\log$ function, 
	\begin{equation*}
	\bar{g}(t) = \log\left(\frac{1}{t_i}\right) \geq \log\left(\frac{1}{t}\right) - \frac{(t_i - t)}{t}.
	\end{equation*}
	As a result,
	\begin{equation*}
	\bar{g}(t) - \frac{g(t)}{2} \geq \frac{\log\left(\frac{1}{t}\right)}{2} - \frac{(t_i - t)}{t} \geq 1/2 - 1/2 = 0.
	\end{equation*}
\end{proof}

The proof of Theorem \ref{thm: inverse_mixing_time_lower_bound} also on a parameter -- which we term \emph{discrete local spread} -- to handle the mixing over very small steps. Formally, the discrete local spread $\pi_1(G)$ is given by
\begin{equation}
\label{eqn: local_spread}
\pi_1(G) := \frac{d_{\min}(G)^2}{10 \vol(V; G)} 
\end{equation}
where $d_{\min}(G) = \min_{v \in V} d(v)$ is the minimum degree in $G$. Intuitively, the discrete local spread gauges how much the walk given by $\Wbf$ has mixed after one step, starting from any node $v$. We will denote $\pi_1(G_{n,r}[\Csig[\Xbf]])$ by $\widetilde{\pi}_{1,n}$.

\begin{lemma}
	\label{lem: local_spread_lb}
	For $\Csig$ satisfying the conditions of Theorem \ref{thm: inverse_mixing_time_lower_bound}:
	\begin{equation*}
	\liminf_{n \to \infty}~ \widetilde{\pi}_{1,n} \geq \frac{\lambda_{\sigma}^2}{\Lambda_{\sigma}^2} \frac{r^d}{(2D)^d}
	\end{equation*}
\end{lemma}

\textcolor{red}{Prove Lemma \ref{lem: local_spread_lb}}.

\subsection{Cut and volume estimates}
\begin{lemma}
	\label{lem: expected_number_boundary_points}
	Under the conditions of Theorem \ref{thm: conductance_upper_bound}, and for any $r < \sigma/2d$,
	\begin{equation*}
	\Pbb(\Csigr) \leq 2 \nu(\Csig) \frac{rd}{\sigma}  \left(\lambda_{\sigma} - \frac{r^{\gamma}}{\gamma + 1}\right)
	\end{equation*}	
\end{lemma}
\begin{proof}
	Recalling that $f$ is the density function for $\Pbb$, we have
	\begin{equation}
	\label{eqn: integral_over_epsilon_neighborhood}
	\Pbb(\Csigr) = \int_{\Csigr} f(x) dx
	\end{equation}
	We partition $\Csigr$ into slices, based on distance from $\Csig$, as follows: for $k \in \N$,
	\begin{equation*}
	\mathcal{T}_{i,k} = \set{x \in \Rd: t_{i,k} < \frac{\rho(x, \Csig)}{r} \leq t_{i+1,k}}, ~~ \Csigr = \bigcup_{i = 0}^{k-1} \mathcal{T}_{i,k}
	\end{equation*}
	where $t_i = i/k$ for $i = 0, \ldots, k - 1$. As a result,
	\begin{equation*}
	\int_{\Csigr} f(x) dx = \sum_{i = 0}^{k-1} \int_{\mathcal{T}_{i,k}} f(x) dx \leq \sum_{i = 0}^{k-1} \nu(\mathcal{T}_{i,k}) \max_{x \in \mathcal{T}_{i,k}} f(x).
	\end{equation*}
	We substitute
	\begin{equation*}
	\nu(\mathcal{T}_{i,k}) = \nu(\Csig + rt_{i+1,k}B) - \nu(\Csig + rt_{i,k}B) := \nu_{i+1,k} - \nu_{i,k}. 
	\end{equation*}
	where for simplicity we've written $\nu_{i,k} = \nu(\Csig + rt_{i+1,k}B)$.
	This, in concert with the upper bound
	\begin{equation*}
	\max_{x \in \mathcal{T}_{i,k}} f(x) \leq \lambda_{\sigma} - (rt_{i,k})^{\gamma},
	\end{equation*}
	which follows from \ref{asmp: bounded_density} and \ref{asmp: low_noise_density}, yields
	\begin{align}
	\label{eqn: telescoping_sum}
	\sum_{i = 0}^{k-1} \nu(\mathcal{T}_{i,k}) \max_{x \in \mathcal{T}_{i,k}} f(x) & \leq \sum_{i = 0}^{k-1} \biggl\{ \nu_{i+1,k} - \nu_{i,k} \biggr\} \biggl( \lambda_{\sigma} - (rt_{i,k})^{\gamma} \biggr) \nonumber \\
	& = \sum_{i = 1}^{k} 
	\underbrace{\nu_{i,k} \biggl( \left[\lambda_{\sigma} - (rt_{i,k})^{\gamma}\right] -  \left[\lambda_{\sigma} - (rt_{i-1,k})^{\gamma}\right]\biggr)}_{:= \Sigma_k} + \underbrace{\biggl(\nu_{k,k}\left[\lambda_{\sigma} - r^{\gamma}\right] - \nu_{1,k}\lambda_{\sigma} \biggr)}_{:= \xi_k}
	\end{align}

@@

\subsection{Proof of Theorem \ref{thm: inverse_mixing_time_lower_bound}}

\textcolor{red}{Give proof structure.}

We begin by introducing some more notation. Take $G = (V,E)$ to be an undirected and unweighted graph, with associated adjacency matrix $\Abf$, random walk matrix $\Wbf$, and stationary distribution $\pibf$. We slightly overload notation and let the \emph{conductance function} be given by
\begin{equation*}
\Phi(t; G) := \min_{\substack{S \subseteq V \\ \pi(S) \leq t} } \Phi(S; G)
\end{equation*}

We will pay special attention to the conductance function evaluated over the subgraph $G_{n,r}[\Csig[\Xbf]]$. For ease of notation, we therefore introduce $\widetilde{\Phi}_{n,r}(t) = \Phi(t; G_{n,r}[\Csig[\Xbf]])$.

Lemma \ref{lem: montenegro} is from the \textcolor{red}{PhD thesis of Montenegro.} It leverages the conductance function to produce an upper bound on the \textcolor{red}{total variation distance} between the random walk 
\begin{equation*}
\rhobf^t = \frac{1}{t} \sum_{s = 1}^{t} e_v \Wbf^s
\end{equation*}
and the stationary distribution $\pibf$.

\begin{lemma}
	\label{lem: montenegro}
	Let $\Wbf$ be the random walk matrix over a graph $G = (V,E)$, with stationary distribution $\pibf$. Then, for any $v \in V$:
	\begin{equation*}
	\norm{\rho^t - \pibf}_{TV} \leq \max\left\{ \frac{1}{4}, \frac{1}{10} +  \frac{70}{t}\left(\frac{4}{\Phi^2(\pi_1(G); G)} + \int_{x = \pi_1(G)}^{1/2} \frac{4}{x \Phi^2(x; G)} dx\right) \right\}
	\end{equation*}
	where $\pi_1(G)$ is defined as in \eqref{eqn: local_spread}.
\end{lemma}

\begin{proof}
	In \textcolor{red}{the PhD thesis of Montenegro}, letting
	\begin{equation*}
	h^t(x_0) = \sup_{S: \pi(S) < x_0}  \bigl(\rhobf^t(S) - \pibf(S) \bigr)
	\end{equation*}
	the following statement is shown: for $\pi_{\min} < x_0 < 1/2$, 
	\begin{equation*}
	\norm{\rho^t - \pibf}_{TV} \leq \max\left\{ \frac{1}{4}, h^t(x_0) +  \frac{70}{t}\left(\frac{4}{\Phi^2(x_0); G)} + \int_{x = x_0}^{1/2} \frac{4}{x \Phi^2(x; G)}\right) \right\}
	\end{equation*}
	It is also observed therein that the function $h^t(x)$ is decreasing for any fixed $x$, as $t$ increases. All that remains for us to show is that $h^s(\pi_1(G)) \leq 1/10$ for some $s \leq t$. We choose $s = 1$, and note that for any set $S \subset V$ with $\pibf(S) \leq \pi_1(G)$,
	\begin{equation*}
	\rhobf^1(S) \leq \frac{\abs{S}}{d_{\min}(G)} \leq \frac{\pibf(S) \vol(V; G)}{d_{\min}(G)^2} \leq \frac{1}{10},
	\end{equation*}
	where we use the fact that $\pibf(S) \geq \frac{\abs{S} d_{\min}(G)}{\vol(V; G)}$.
\end{proof}

We turn to lower bounding $\widetilde{\Phi}_{n,r}(t)$. First, we exhibit a lower bound on a continuous space analogue, over the set $\Csig$. Let $\nu_{\Pbb}(\cdot)$ denote the weighted volume; formally, for $\Sset \subset \Rd$
\begin{equation*}
\nu_{\Pbb}(\Sset) := \int_{\Sset} f(x) dx
\end{equation*}

The $r$-ball walk over $\Csig$ is a Markov chain with transition probability for $x \in \Csig, \Sset, \Sset' \subset \Csig$ given by,
\begin{equation*}
P_{\Pbb, r}(x;\Sset) := \frac{\nu_\Pbb(\Sset \cap B(x,r))}{\nu_\Pbb(\Csig \cap B(x,r))}, ~~~ Q_{\Pbb, r}(\Sset, \Sset') := \int_{x \in \Sset} f(x) P_{\Pbb, r}(x;\Sset') dx. 
\end{equation*}
stationary distribution defined by
\begin{equation*}
\ell_{\Pbb,r}(x) := \frac{\nu_\Pbb(\Csig \cap B(x,r))}{\nu_{\Pbb}(B(x,r))}, ~~~ \pi_{\Pbb,r}(\Sset) := \frac{1}{\int_{\Csig} f(x) \ell_{\Pbb,r}(x) dx} \int_{\Sset} f(x) \ell_{\Pbb,r}(x) dx
\end{equation*}
and corresponding conductance function
\begin{equation*}
\widetilde{\Phi}_{\Pbb,r}(t) := \min_{\substack{\Sset \subset \Csig, \\ \pi_{\Pbb,r}(\Sset) \leq t} } \frac{Q_{\Pbb,r}(\Sset, \Csig \setminus \Sset)}{\pi_{\Pbb,r}(\Sset)}
\end{equation*}
Lemma \ref{lem: continuous_conductance_lb} (\textcolor{red}{a weighted analogue of Theorem 4.6 of Kannan}) provides a lower bound on $\widetilde{\Phi}_{\Pbb, r}(t)$, as well as a stepwise approximation to $\widetilde{\Phi}_{\Pbb, r}$.
\begin{lemma}
	\label{lem: continuous_conductance_lb}
	Under the conditions on $\Csig$ given by Theorem \ref{thm: inverse_mixing_time_lower_bound}, the following bounds hold: 
	\begin{itemize}
		\item 
		for $0 < t < 1/2$,
		\begin{equation*}
		\widetilde{\Phi}_{\Pbb,r}(t) > \min\left\{\frac{1}{288\sqrt{d}},\frac{r}{81 \sqrt{d}D}\log\left(\frac{\lambda_{\sigma}^2}{\Lambda_{\sigma}^2 t}\right)\right\} \cdot \frac{\lambda_{\sigma}^4}{\Lambda_{\sigma}^4}
		\end{equation*}
		\item
		Let
		\begin{equation*}
		m = \frac{2^{d+1}D^d \Lambda_{\sigma}^2}{r^d \lambda_{\sigma}^2}
		\end{equation*}
		 and $t_i = (i + 1)/m$ for $i = 0, \ldots, m - 1$. Then, for $1/m < t < 1/2$
		\begin{equation*}
		\overline{\Phi}_{\Pbb,r}(t) > \min\left\{\frac{1}{288\sqrt{d}},\frac{r}{162 \sqrt{d}D}\log\left(\frac{\Lambda_{\sigma}^2}{\lambda_{\sigma}^2 t}\right)\right\} \cdot \frac{\lambda_{\sigma}^4}{\Lambda_{\sigma}^4}
		\end{equation*}
		where $\overline{\Phi}_{\Pbb,r}(t)$ is defined as in \eqref{eqn: stepwise_approximation} with respect to $t_0, \ldots t_{m - 1}$. 
	\end{itemize}

\end{lemma}
\begin{proof}
	We begin by stating directly the result of \textcolor{red}{Kannan et al.}: for $0 < t < 1/2$, and any $S$ such that $\pi_{\nu,r}(S) = t$,
	\begin{equation}
	\label{eqn: kannan_1}
	\frac{Q_{\nu,r}(\Sset, \Csig \setminus \Sset)}{t} > \min\left\{\frac{1}{288\sqrt{d}},\frac{r}{81 \sqrt{d}D}\log\left(\frac{1}{t}\right)\right\}.
	\end{equation}
	Now, we note that
	\begin{equation*}
	\pi_{\Pbb,r}(S) \leq \pi_{\nu,r}(S) \cdot \frac{\Lambda_{\sigma}^2}{\lambda_{\sigma}^2}, ~~~ Q_{\Pbb,r}(\Sset, \Csig \setminus \Sset) \geq Q_{\nu,r}(\Sset, \Csig \setminus \Sset) \cdot \frac{\lambda_{\sigma}^2}{\Lambda_{\sigma}^2}
	\end{equation*}
	and plugging these estimates in to \eqref{eqn: kannan_1} gives the final bound on $\widetilde{\Phi}_{\Pbb,r}(t)$.  The bound on $\overline{\Phi}_{\Pbb,r}(t)$ then follows immediately from application of Lemma \ref{lem: stepwise_approximation}.
\end{proof}

The introduction of the stepwise approximation allows us to make use of Lemma \ref{lem: consistency_of_conductance_function}, which gives us (pointwise) consistency of the discrete graph functionals $\widetilde{\Phi}_{n,r}(t)$ to the continuous functionals $\widetilde{\Phi}_{\Pbb,r}(t)$.

\begin{lemma}
	\label{lem: consistency_of_conductance_function}
	Under the conditions on $\Csig$ given by Theorem \ref{thm: inverse_mixing_time_lower_bound}, fix any $0 < t < 1/2$. Then the following statement holds: with probability one, as $n \to \infty$,
	\begin{equation*}
	\liminf_{n \to \infty} \widetilde{\Phi}_{n,r}(t) \geq \widetilde{\Phi}_{\Pbb,r}(t)
	\end{equation*}
	As a consequence, $m$ and $(t_i)_{i=1}^{m}$ defined as in Lemma \ref{lem: continuous_conductance_lb}, we have that
	\begin{equation}
	\label{eqn: consistency_of_stepapprox_conductance_function}
	\liminf_{n \to \infty} \widetilde{\Phi}_{n,r} \geq \overline{\Phi}_{\Pbb,r}
	\end{equation}
\end{lemma}
Lemma \ref{lem: consistency_of_conductance_function} stems from \textcolor{red}{Garcia Trillos 16}, with a pair of notable distinctions: here, we do not allow the radius $r$ to go to zero, but rather set it to be constant, and in the minimization we have an additional constraint on the measure $\pi(\cdot)$, in both the discrete and continuous functionals. As such, we will need to re-prove a number of the statements of \textcolor{red}{Garcia Trillos 16} to work in this context. We defer this work to Section \ref{subsection: proof_of_pointwise consitency_of_conductance_function}, and here will only show that  \eqref{eqn: consistency_of_stepapprox_conductance_function} is implied immediately by the pointwise result.

\begin{proof}[Proof of \eqref{eqn: consistency_of_stepapprox_conductance_function}]
We take as given that for any $0 < t < 1/2$,
\begin{equation*}
\lim_{n \to \infty} \widetilde{\Phi}_{n,r}(t) \geq \widetilde{\Phi}_{\Pbb,r}(t).
\end{equation*}
In particular, this will occur for $t_0, t_1, \ldots, t_m$ and therefore
\begin{equation*}
\lim_{n \to \infty} \overline{\Phi}_{n,r} \geq \overline{\Phi}_{\Pbb,r}
\end{equation*}
uniformly over $[1/m,1/2]$. But by Lemma \ref{lem: stepwise_approximation}, $\widetilde{\Phi}_{n,r} \geq \overline{\Phi}_{n,r}$, so we have shown \eqref{eqn: consistency_of_stepapprox_conductance_function}.
\end{proof}

Lemma \ref{lem: consistency_of_conductance_function} coupled with Lemma \ref{lem: montenegro} yields a bound on the total variation mixing time.
\begin{lemma}

\end{lemma}

\subsection{Proof of pointwise consistency of conductance function.}
\label{subsection: proof_of_pointwise consitency_of_conductance_function}
